---
name: "Aspire Data Solutions"
title: "Applying Bibliometric analysis and text mining to COVID-19 publications?"
author: Mihiretu Kebede(PhD)
url: http://www.mihiretukebede.com/ 
google_analytics: "UA-173027539-1"
date: "2020-08-04`"
description: |
  Let's have a quick overview of what on COVID-19 is just indexed in PubMed only in August 03/2020.

collections:
  posts:
      share: [twitter, linkedin, facebook, google-plus, pinterest]
      
output:
  distill::distill_article:
    self_contained: false
---

<link rel="stylesheet" href="styles.css" type="text/css">

#Plan of attack

## Identify source of data: copy this and search on PubMed. 
- Let's do a quick PuBmed search: (covid 19) AND (("2020/08/02"[Date - Publication] : "3000"[Date - Publication]))

- With the above search terms, we have 601 COVID-19 related records were indexed in PubMed 
- ALternatively you can find the search link [here](https://pubmed.ncbi.nlm.nih.gov/?term=%28covid+19%29+AND+%28%28%222020%2F07%2F31%22%5BDate+-+Publication%5D+%3A+%223000%22%5BDate+-+Publication%5D%29%29).

- We can now open these records using our citation manager. 
- Since the file is in RIS format, we must convert it to BibTex so that we can be be able to benefit from **Bibliomterix** package. 
- I will come back in the future in another blog post to show you on how to convert RIS file to BibText or from CSV/XSL to EndNote-readable format. For today, let's just use the converted [COVID19.bib](https://github.com/Mihiretukebede/mihiretukebede.gitlub.io/blob/master/posts/2020-08-03-2020-08-03-covid19/covid19.bib)


## iniatilize
Set working directory

### Load necessary packages

```{r, cache=T, include=T, echo=T, message=FALSE, warning=FALSE}
library(bibliometrix) #for bibliometric analysis
library(dplyr) #for data management
library(ggplot2) #for plotting
library(tidytext) #for text mining
library(bib2df) #for converting bib file to data frame
```

# Load data
In another blog post, I will come back to show you how to convert RIS reference file to BibTex of csv file to EndNote readable file formats. For now, let's first use *bibliometrix* and then *bib2df* package to convert our [covid bib file](https://github.com/Mihiretukebede/mihiretukebede.gitlub.io/blob/master/posts/2020-08-03-2020-08-03-covid19/covid19.bib) to data frame. 


# Bibliometrics analysis using the biblimotrix package

```{r, cache=T, include=T, echo=T, message=FALSE, warning=FALSE}
covid19_bibanalysis <- convert2df("covid19.bib", dbsource = "isi", format = "bibtex")

#As usual, isnpect the data
dim(covid19_bibanalysis) #601 records, 23 variables
glimpse(covid19_bibanalysis)

```

# Now bibliometrics analysis

```{r, cache=T, include=T, echo=T, message=FALSE, warning=FALSE}
results <- biblioAnalysis(covid19_bibanalysis, sep = ",")
options(width=100)
s <- summary(object=results, k=10, pase=FALSE)
plot(x=results, k=10, pause=FALSE)

#since all of them are indexed in August 2020, this may not be useful analysis. 

#Let's see sankey plots. But, before that we must remove missing values
threeFieldsPlot(covid19_bibanalysis, fields = c("AU", "DE", "SO"))

#Look at this plot. It maps Authors with keywords and with the journals. With 61 articles indexed in PubMed in one day, a journal called, EMERGING MICROBES \\& INFECTIONS  is leading the league


## Now let's do some clustering
cs <- conceptualStructure(covid19_bibanalysis, field="AB", 
                          method="CA", minDegree = 4, 
                          stemming=FALSE, labelsize = 10, documents=10)
```

# Now, let's use bib2df

It is also possible to continue using the `covid19_bibanalysis`data. However, it is also good to try another package, `bib2df`. We will use this data frame for our text mining analysis. Sometimes, the data frame converted using *bibliometrix*  package may not also work properly with *dplyr*.  

```{r, cache=T, include=T, echo=T, message=FALSE, warning=FALSE}
library(bib2df)
covid19 <- bib2df("covid19.bib")
```


# Inspect the data

```{r, cache=T, include=T, echo=T, message=FALSE, warning=FALSE}
glimpse(covid19) 
#Let's choose only some of our variabees

covid19data <- covid19 %>% 
  select("AUTHOR", "TITLE", "KEYWORDS", "ABSTRACT", "JOURNAL","DOI")

dim(covid19) #601 records, 23 variables
which(!complete.cases(covid19$DOI)) # All of them have missing values
sum(is.na(covid19$ABSTRACT)) #159 studies have no abstract
sum(is.na(covid19$TITLE)) # 0 records have missing values
sum(is.na(covid19$DOI)) #2 studies have no doi
sum(!is.na(covid19$ABSTRACT)) #442 records have abstracts

#Let's filter the data to retreive only the records with Abstract(AB)

```

```{r, cache=T, include=T, echo=T, message=FALSE, warning=FALSE}
#Let's filter the data to retreive only the records with Abstract(AB)
#As we have seen above, we have 159 records without abstracts
#use dplyr to filter
library(dplyr)
covid19new <- covid19 %>%
  filter(!is.na(ABSTRACT)) #remove all records with missing abstracts
sum(!is.na(covid19new$ABSTRACT)) #442 records have abstracts
   #BINGO, we excluded the studies without Abstract. Now we, can play with text mining. 

```

# Text mining
A work in progress!



# Contact 
Please mention [MihiretuKebede1](https://www.https://twitter.com/MihiretuKebede1) if you tweet this post. 


