---
name: "Aspire Data Solutions"
output_dir: "."
title: "Vizualizing the prevalence of diabetes in six European countries"
author: Mihiretu Kebede (PhD)
url: http://www.mihiretukebede.com/ 
google_analytics: "UA-173027539-1"
date: "2020-07-25`"
description: |
  This is a quick demonstration of diabetes prevalence in six European countries. Tha data are from the IHME. 
  
collections:
  posts:
      share: [twitter, linkedin, facebook, google-plus, pinterest]
      
output:
  distill::distill_article:
    self_contained: false
---


<link rel="stylesheet" href="styles.css" type="text/css">

# A step by-step guide on how to improve a simple scatter plot 

- If you would like to reproduce these codes, you can download the data from my github repository. 
- The file name is Eurodiabetes.csv or simply click  [here](https://github.com/Mihiretukebede/mihiretukebede.gitlub.io/tree/master/Eurodiabetes). If you want to access the codes, click [this](https://github.com/Mihiretukebede/mihiretukebede.gitlub.io/blob/master/Eurodiabetes/Europediabetesanimation.R). 

- Once you download the data, don't forget to set your working directory! 

# My R version
- I used the following R version to produce this result. 

```{r, include=T, echo=TRUE}
version
```

# Required packages
```{r, include=T, echo=FALSE}
library(readr) #to read csv file
library(dplyr) #for data manipulaion 
library(ggplot2) # for awesome plotting
library(gganimate) #for animating ggplot objects
library(scales) # or customizing axis 
library(lattice) #for enhancing graphics
library(directlabels) #for directly labeling!
```

## Let's load the data and have a closer look at it. 

```{r, include=T, echo=FALSE}
# The data is for all countries included in GBD studies
diabetes <- read_csv("Eurodiabetes.csv") 
dim(diabetes) #980 observations and 9 variables. 
str(diabetes) 
```

## Let's choose only 5 European countries with high diabetes prevalence

```{r, include=TRUE, echo=FALSE}
ger_au_ch <- diabetes %>% 
  filter(Location %in% c("Austria", "Germany", "Switzerland", "Denmark", "Portugal", "Finland")) 
ger_au_ch <- na.omit(ger_au_ch) #Remove missing values
ger_au_ch$Prev <- ger_au_ch$Value*100 #Prevalence in percent. 
str(ger_au_ch) #structure of the data, variable types
dim(ger_au_ch) #168 rows, 10 columns 

is.factor(ger_au_ch$Year) #check if Year is saved as factor variable
ger_au_ch$yearfactor <- factor(ger_au_ch$Year) #convert it to factor and save it as Yearfactor

ger_au_ch$Yearnumeric <- as.numeric(ger_au_ch$Year) #change it to numeric and save it as Year Numeric
```
# Plot
Since we already have everything we need for plotting, we can start using ggplot2

```{r, layout="l-body-outset", preview=TRUE, include=TRUE, echo=FALSE}
plot <- ggplot(ger_au_ch, aes(x=Yearnumeric, y=Prev, col=Location)) + 
  geom_line()
plot
```

### Now we can start making it more advanced

```{r, include=TRUE, echo=FALSE}
euro_anim <- ggplot(ger_au_ch, aes(x=Yearnumeric, y=Prev, col=Location)) + 
  geom_point(size=6) + transition_time(Yearnumeric)  +
  shadow_mark()  +  
  scale_x_continuous(name ="Year",
                     breaks= c(1990,1995,2000,2005, 
                               2010, 2015, 2020)) +
  
  xlab("Year") +
  ylab("Prevalence of diabetes in %") +
  labs(col="Country") + 
  ggtitle("Prevalence of diabetes cases in six European countries") +
  
  theme(
    axis.title.x = element_text(color = "Blue", size=15),
    axis.title.y = element_text(color = "Blue", size=15),
    axis.text.x = element_text(size = 15),
    axis.text.y = element_text(size = 15),
    
    plot.title = element_text(size=20),
    
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 10),
    legend.position = "None",
    
    text = element_text(family = "Comics Sans MS")
  ) + ease_aes('cubic-in-out') +
  geom_dl(aes(label=Location),
          method=list("last.points",rot=40)) 

```

## Now we can use gifski_renderer to loop or stop looping the gif 

```{r, echo=FALSE, include=TRUE}
animate(euro_anim, renderer = gifski_renderer(loop = F), width = 700, height = 700, duration = 15) # when you assign loop=TRUE, the gif starts playing again 
```


# Let's tweak few things and see what happens
```{r, echo=FALSE, include=TRUE}

euro_anim <- ggplot(ger_au_ch, aes(x=Yearnumeric, y=Prev, col=Location)) + 
  geom_point(size=6) + transition_time(Yearnumeric)  +
  shadow_mark()  +  
  scale_x_continuous(name ="Year",
                     breaks= c(1990,1995,2000,2005, 
                               2010, 2015, 2020)) +
  
  xlab("Year") +
  ylab("Prevalence of diabetes in %") +
  labs(col="Country") + 
  ggtitle("Prevalence of diabetes cases in Austria, Germany and Switzerland") +
  shadow_wake(wake_length = 0.1, alpha = FALSE) +
  theme(
    axis.title.x = element_text(color = "Blue", size=15),
    axis.title.y = element_text(color = "Blue", size=15),
    axis.text.x = element_text(size = 15),
    axis.text.y = element_text(size = 15),
    
    plot.title = element_text(size=20),
    
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 10),
    legend.position = "None",
    
    text = element_text(family = "Comics Sans MS")
  ) + ease_aes('cubic-in-out') +
  geom_dl(aes(label=Location),
          method=list("last.points",rot=40)) 

animate(euro_anim, renderer = gifski_renderer(loop = F), width = 700, height = 700, duration = 15)

```

```{r, include=TRUE, echo=FALSE}
euro_anim3 <- ger_au_ch %>% 
  ggplot( aes(x=Yearnumeric, y=Prev, col=Location)) + geom_line() + geom_line() +
  geom_point() +
    transition_reveal(Yearnumeric) + 
  shadow_mark()  +  
  scale_x_continuous(name ="Year",
                     breaks= c(1990,1995,2000,2005, 
                               2010, 2015, 2020)) +
  xlab("Year") +
  ylab("Prevalence of diabetes in %") +
  labs(col="Country") + 
  ggtitle("Prevalence of diabetes cases in six European countries") +
  shadow_wake(wake_length = 0.1, alpha = FALSE) +
  theme(
    axis.title.x = element_text(color = "Blue", size=15),
    axis.title.y = element_text(color = "Blue", size=15),
    axis.text.x = element_text(size = 15),
    axis.text.y = element_text(size = 15),
    
    plot.title = element_text(size=20),
    
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 10),
    legend.position = "None",
    
    text = element_text(family = "Comics Sans MS")
  ) + ease_aes('cubic-in-out') +
  geom_dl(aes(label=Location),
          method=list("last.points",rot=40))

euro_anim3  
```


That is it. I hope, you like it. See you in my next post.

## Contact 
Please mention [MihiretuKebede1](https://www.https://twitter.com/MihiretuKebede1) if you tweet this post. 

