---
name: "Aspire Data Solutions"
output_dir: "."
title: "Animating scatter plots"
author: Mihiretu Kebede
url: http://www.mihiretukebede.com/ 
preview: F:/github/githubwebsite/scatterplot.JPG
google_analytics: "UA-173027539-1"
date: "2020-07-21"
description: |
  Creating `Gif` for scatter plots

collections:
  posts:
    disqus: 
      shortname: "animating-scatter-plots"
      hidden: false
      share: [twitter, linkedin, facebook, google-plus, pinterest]
output:
  distill::distill_article:
    self_contained: false
---

## Animating a scatter plot.

`Gif` is a fancy way of data vizualization. It facilitates communication, makes your audience active. 

1. What do I need to do to produce gif? 
First load the packages we need. 
If you don't have these packages install them. 

## Load the data 

```{r, echo=TRUE}
library(readxl)
type2pre <- read_excel("F:/github/githubwebsite/type2GBDcompare.xls")
View(type2pre)
```

## Source of data 
The data is about prevalence of `type 2 diabetes` for 20 countries from 1990-2017

```{r, include=TRUE, echo=TRUE}
# You can downloaded such data from the IHME GBD compare websisite. You can select any indicator(prevalence, DALYs, mortality, etc). Check their website [GBD](https://vizhub.healthdata.org/gbd-compare/) 

#We will plot prevalence 
is.factor(type2pre$Year) #to check whether year is coded as numeric variable
type2pre$yearfactor <- factor(type2pre$Year) #To save year as factor variable
type2pre$Yearnum <- as.numeric(type2pre$Year) #To save year as numeric variable
```

## Install the necessary packages 
To install the latest version of the packages 
`library(devtools)` 

- NB if you don't have the necessary packages, you need to install them with the following codes

`devtools::install_github("thomasp85/gganimate")`
`devtools::install_github("thomasp85/transformr")` transformr is a dependency that helps you with transitions for polygons and lines. 

## Load the necessary packages

```{r, echo=TRUE, include=TRUE}
library(ggplot2) # for plotting 
library(gganimate) #for animating your plot
library(scales) # for scaling your x or y-axis 

```

## The plot
```{r setup, include=TRUE}
#Basic ggplot
myplot <- ggplot(type2pre, aes(x=Yearnum, y=Prevalence)) + 
  geom_point(aes(color=Location, size=3))

myplot

## the gganimate part of code
animateplot <- myplot + transition_time(Yearnum) +
  shadow_mark() + scale_x_continuous(limits = c(1990,2017)) + 
  xlab("Year") + ylab("Prevalence per 100,000")

##NB: The shadow_mark() function is to include the previous plots in the scatter plot.Then each point moves across the variable that passess through the transition_time() function
## the animation is done by passing the ggplot on the animate function of gganimate
animate(animateplot, width = 700, height = 500)

```

*There are a lot you can do with `gganimate`. Google for more!

## References 
1. https://github.com/thomasp85/gganimate#old-api
2. https://paldhous.github.io/ucb/2018/dataviz/week14.html 

## Contact 
 [@MihiretuKebede1](https://twitter.com/MihiretuKebede1) 