---
name: "Aspire Data Solutions"
title: "Public reactions to Abiy Ahmed's Facebook posts"
author: Mihiretu Kebede(PhD)
url: http://www.mihiretukebede.com/ 
google_analytics: "UA-173027539-1"
date: "2020-08-04`"
description: |
  Let'S have some fun in visualizing Facebook reactions to Abiy Ahme's Facebook posts from July 27 to August 04.

collections:
  posts:
      share: [twitter, linkedin, facebook, google-plus, pinterest]
      
output:
  distill::distill_article:
    self_contained: false
    widerscreen: true
---


# Disclaimer
This blog post is simply about visualizing the data. It doesn't reflect the political affiliation or opinion of the author.  

# Data
I have manually collected Prime Minister Abiy Ahmed's one month Facebook posts (From June 27 until August 24). I collected simply three variables: date Abiy Ahmed posted on Facebook, type of Facebook reaction (eg: Like, LOve, angry, sad, etc...) and number of reactions. 

# Initialize
# Required packages 
```{r, cache=T, include=T, echo=T, message=FALSE, warning=FALSE}
library(readxl) #to read the excel data
library(ggplot2) #for awesome plotting
library(directlabels) #for labelling directly on the plots
```

``````{r, cache=T, include=T, echo=T, message=FALSE, warning=FALSE}
library(readxl)
abiy_ahmed <- read_excel("abiy_ahmed.xlsx")
# As usual 
dim(abiy_ahmed)
glimpse(abiy_ahmed)
```

# Plot
``````{r, cache=T, include=T, echo=T, message=FALSE, warning=FALSE}
plot <- ggplot(abiy_ahmed, aes(x=Date_posted, y=Count, col=Reaction)) + geom_point() +
  geom_line()
direct.label(plot, "last.qp")
```


# Contact 
Please mention [MihiretuKebede1](https://twitter.com/MihiretuKebede1) if you tweet this post. 

