---
title: "Classifying abstracts using regularized models: covid-19 or sars-2003?"
theme: theme.css
author: Mihiretu Kebede(PhD)
url: http://www.mihiretukebede.com/ 
google_analytics: "UA-173027539-1" 

description: |
   Classifying abstracts using regularized models: covid-19 or sars-2003? 
collections:
  posts:
      share: [twitter, linkedin, facebook, google-plus, pinterest]
      
output:
  distill::distill_article:
    self_contained: false
    widerscreen: true
---

# Introduction

Classifying documents to a certain list of categories can provide valuable insights and possibly make the documents more manageable. Document classification is one of the application areas of machine learning. Natural Language Processing(NLP) and machine learning methods can be used to automate classification of documents such as emails(ham/spam categories), articles, books, response to survey questions, sentiments, product reviews(negative/positive), etc.

There are handful of very good algorithms that can automagically handle text data. In this blog post, I would like to experiment regularized logistic regression machine learning models for classifying abstracts.

I am interested in predicting abstracts that will fall into "SARS_2003" and "COVID_19" categories. For this task, I will use the same data sets that I have used in my previous blog posts. I saved the references files in excel from my previous post. I will simply load this excel file and play with the abstract using `qunateda` and `glmnet` packages. 😍

# Data

I load the data and see a summary of number of characters used to write the abstracts. For this analysis, I will use only abstracts having 200 characters. I will just filter out the abstracts on the lower side of the extreme.

```{r setup, include=FALSE}
library(readxl)
library(dplyr)
covid_sars <- read_excel("covidsars.xlsx")
glimpse(covid_sars)

# Number of characters
covid_sars$abstract_length <- nchar(covid_sars$abstract)
summary(covid_sars$abstract_length) 

# For this analysis, I will use only abstracts that are written with more than 200 characters
covid_sars <- covid_sars %>% 
  filter(abstract_length > 200)
```

# Check number of covid/sars papers

The covid abstracts were originally labeled as "included" and sars abstracts as "excluded". I will simply create a category variable and change the labels to "covid" or "sars". About 52% of the abstracts have "covid" labels the remaining is "sars".

```{r, cache=T, include=T, echo=T, message=FALSE, warning=FALSE}

covid_sars$category <- covid_sars$Include
covid_sars$category[covid_sars$category=="included"]<-"covid"
covid_sars$category[covid_sars$category=="not included"]<-"sars"

table(covid_sars$category)
round(table(covid_sars$category) %>% prop.table()*100, 1)

```

# Split the data

Before we do any steps in a text analysis, it is recommended to split the data. Splitting the data after tokenization is not a good approach. So, we will instead split the data into train and test set, tokenize the train and test sets separately, build models, match the variables of the test data set with the train. Finally, we will predict the test data and then evaluate our predictions.

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
library(caret) #For splitting 

set.seed(1234)

trainIndex <- createDataPartition(covid_sars$category, p = .8, 
                                  list = FALSE, 
                                  times = 1)

train <- covid_sars[trainIndex,]
test  <- covid_sars[-trainIndex,]

table(train$category)
table(test$category)
nrow(train)
```

# Text analysis

If you are interested in text analysis, I reccommend visiting Quanteda website. Quanteda is a great package for text analysis. One of the great advantage of quanteda is it is super fast, have so many powerful functions. I have tried five different R packages for text data. Based on my personal taste, I would rank them as follows: Quanteda, txt2vec, tidytext, tm.

## Create Corpus

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
require(quanteda)

#Train data
tr <- train[, 6] # the sixth variable is unique label. I will use it as identifier. 
traincorpus <- corpus(train$abstract,
                      docvars = data.frame(trainvars=names(tr)))
#Test data
ts <- test[, 6]

testcorpus <- corpus(test$abstract,
                     docvars = data.frame(testvars=names(ts)))
summary(traincorpus,2)

# Connect the labels with the corpuses
docid_train <- train$label

docnames(traincorpus) <- docid_train
head(traincorpus,1)

docid_test <- test$label
docnames(testcorpus) <- docid_test

summary(testcorpus, 2)
summary(traincorpus, 4)
```

# Tokenize

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
traintokens <- tokens(traincorpus,
                      remove_punct = TRUE,
                      remove_url = TRUE,
                      remove_numbers = TRUE)

traintokens <- tokens_remove(traintokens, 
                             pattern=stopwords('en'))

testtokens <- tokens(testcorpus,
                     remove_punct = TRUE,
                     remove_url = TRUE,
                     remove_numbers = TRUE)

testtokens <- tokens_remove(testtokens, 
                             pattern=stopwords('en'))

```

# Construct the DFM objects

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
dfmat_train <- dfm(traintokens)

dfmat_test <- dfm(testtokens)

head(dfmat_train,2)
head(dfmat_test,2)
```

The training data has 12,119 features and is 99.2% sparse, while the test data has 5,484 features and 97% sparsity. I will not do anything to reduce the sparsity. But, you may have to do it if you have a large number of observations. Quanteda's dfm_trim() can do that for you.

TF-IDF weighting is known to improve prediction performance. I will use that here too.

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
dfmat_train_tfidf <- dfm_tfidf(dfmat_train)
dfmat_test_tfidf <- dfm_tfidf(dfmat_test)
```

Let's inspect the two tfidf data that were created above.

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
head(dfmat_train_tfidf, 2)
head(dfmat_test_tfidf,2)
```

# Model building

## Why not logistic regression?

My data has two class labels(covid vs sars) and all numerical features. Why not logistic regression, then? Well, linear models provide great approaches to predictive modeling given that the assumptions are met! These assumptions (for example: hemoscidascity of variance) are violated when we have more number of features than observations (For example in genetics studies and text analysis, this is often the case). Applying linear models to such data results in biased coefficients, weaker prediction performance scores, overfitting or high out of sample prediction error problems. Hence, penalizing the estimated model coefficients was devised. This method of penalizing linear models is called "Regularization". There are three most commonly used approaches to regularization for logistic regression: Ridge, LASSO, and Elastic net.

In Ridge penalty, the estimated model coefficients are penalized by adding the following parameter. SSE $$\begin{equation}
\ SSE = \sum^n_{i=1} \left(y_i - \hat{y}_i\right)^2 
\end{equation}$$

$$\begin{equation} \ SSE + \lambda \sum^p_{j=1} \beta_j^2 \
\end{equation}$$

This is called L\^2 norm. From the above equation if lambda equals 0, the model will be equal to the ordinary least squares model. As approaches to infinity, the penalty will force the model coefficients to be closer to zero but not completely to 0. Ridge penalty is known to systematically handling highly correlated features. In Lasso penalty the model coefficients are penalized by a L1 norm as follows. $$\begin{equation}SSE + \lambda \sum^p_{j=1} | \beta_j |
\end{equation}$$

Lasso penalty unlike ridge pushes all the coefficients all the way to zero. The advantage of Lasso is it improves model performance while also automating feature selection. Only the features that are important will be retianed on the final model.

Elastic net combines both Lasso and Ridge penalty parameters. Elastic net takes advantages of both Lasso and Ridge penalty: effective regularization by automated feature selection as well as effectively handling correlated features.

# Implementations

Probably the most popular package to implement regualrized models is the `glmnet` package. This package is lightening fast to tune cross validated models. I watched one very nice webinar tutorial from Dr Trevor Hastie(one of the authors of this package). He mentioned that it is fast because it is programmed in Fortran. I invite you to watch that great [webinar here](https://www.youtube.com/watch?v=BU2gjoLPfDc&t=2592s&ab_channel=ProfTrevorHastie) I head I heard there are also other packages like `H2O` and `elastic net`. I have never tried any of them.

For regularized models, we have two main tuning parameters: alpha and lambda. In ridge and Lasso the lambda is the only tuning parameter but alpha is set to be 0 and 1, respectively. For tuning lambda, the cv.glmnet() function provides 100 different data driven lambda values and there is no need to do anything else. Since elastic net combines both Lasso and Ridge penalty, we will have two tuning parameters: alpha and lambda. Alpha can take a number values between between 0 and 1, while lambda can have 100 different data driven lambda values by just using `cv.glmnet()` function.

All models require the data to be in a matrix form. The good thing with quanteda is the document feature matrix is already a matrix object we don't need to change the structure of our data. Notice the time required to tune the cross-validated algorithms.

# Ridge

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
# Ridge regression
library(glmnet)
ridge_1 <- glmnet(x = dfmat_train, y = train$category, 
                    alpha = 0, family = "binomial")
#tfidf
ridge_1_tfidf <- glmnet(x = dfmat_train_tfidf, y = train$category, 
                    alpha = 0, family = "binomial")

par(mfrow = c(1, 2))
plot(ridge_1, xvar="lambda", main="Ridge penalty\n\n")
plot(ridge_1_tfidf, xvar="lambda", main="Ridge penalty tfidf\n\n")

```

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
x <- Sys.time()

set.seed(123)
ridge_min <- cv.glmnet(x=dfmat_train,
                   y=train$category,
                   family="binomial", 
                   alpha=0,  # alpha = 0 for ridge regression
                   parallel=TRUE, 
                   intercept=TRUE)

```

Again using the tf-idf weighted data

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
set.seed(123)
ridge_min_tfidf <- cv.glmnet(x=dfmat_train_tfidf,
                   y=train$category,
                   family="binomial", 
                   alpha=0,  # alpha = 0 for ridge regression
                   parallel=TRUE, 
                   intercept=TRUE)

par(mfrow = c(1, 2))
plot(ridge_min, main="Ridge penalty\n\n")
plot(ridge_min_tfidf, main="Ridge penalty_tfidf\n\n")
Sys.time() - x
```

Let's plot the results

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
par(mfrow = c(1, 2))
plot(ridge_1, xvar = "lambda", main = "Ridge penalty\n\n") 
abline(v=log(ridge_min$lambda.min), col = "red", lty = "dashed")
abline(v=log(ridge_min$lambda.1se), col = "blue", lty = "dashed")

plot(ridge_1_tfidf, xvar = "lambda", main = "Ridge penalty tfidf\n\n") 
abline(v=log(ridge_min_tfidf$lambda.min), col = "red", lty = "dashed")
abline(v=log(ridge_min_tfidf$lambda.1se), col = "blue", lty = "dashed")
```

# Predict the test data sets

Before we predict the test data, we need to do one very key step. We will predict the test data based on the data that the model was trained. So, features on the test data should match with the features on the training data. Otherwise, the prediction will not work. The model cannot understand anything outside of the features that were in the training data. This is very key step in text prediction. Quanteda provides a nice function for that: dfm_match(). It subsets the features of the test data that were part of the training data.

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
dfmat_matched <- dfm_match(dfmat_test, 
                           features = featnames(dfmat_train))
# Match the tfi-idf
dfmat_matched_tfidf <- dfm_match(dfmat_test_tfidf, 
                                 features = featnames(dfmat_train_tfidf))

```

For prediction, I will use the best model from the cross validated models. The best model lies between the model having the minimum lamda value and the model that has a lambda value within 1 se. Here, I will use the minimum lamda value.

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
# Predict 
ridge_min
ridge_min$lambda.1se
ridge_min$lambda.min
actual_class <- as.factor(test$category)
predicted_class.ridge <- predict(ridge_min, newx=dfmat_matched,s="lambda.min", type="class")

tab_class.ridge <- table(predicted_class.ridge, actual_class)

confusionmatrix_ridge <- confusionMatrix(tab_class.ridge, mode="everything", positive="covid")


##tfidf
ridge_min_tfidf
ridge_min_tfidf$lambda.1se
ridge_min_tfidf$lambda.min
actual_class_tfidf <- as.factor(test$category)
predicted_class.ridge_tfidf <- predict(ridge_min_tfidf, newx=dfmat_matched_tfidf, s="lambda.min", type="class")

tab_class.ridge_tfidf <- table(predicted_class.ridge_tfidf, actual_class)

confusionmatrix_ridge_tfidf <- confusionMatrix(tab_class.ridge_tfidf, mode="everything", positive="covid")

confusionmatrix_ridge
confusionmatrix_ridge_tfidf

```

See, all features are retained in the final models. Looking at the prediction performances of the two, we can see that the tfidf-weighted data has better performance. We will try Lasso if we can improve that.

# Lasso penalty

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
## Lasso model
lasso_1 <- glmnet(x = dfmat_train, y = train$category, 
                    alpha = 1, family = "binomial", type.measure="class") 

lasso_1_tfidf <- glmnet(x = dfmat_train, y = train$category, 
                    alpha = 1, family = "binomial", type.measure="class") 

par(mfrow=c(1,2))

plot(lasso_1, xvar="lambda", main="Lasso penalty\n\n")
plot(lasso_1_tfidf, xvar="lambda", main="Lasso penalty tfidf\n\n")

```

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}

x <- Sys.time()
#registerDoMC(cores=2) # parallelize to speed up
set.seed(123)
lasso <- cv.glmnet(x=dfmat_train,
                   y=train$category,
                   family="binomial", 
                   alpha=1,  # alpha = 1: LASSO
                   parallel=TRUE, nfolds = 10,
                   intercept=TRUE) 

# tfidf
set.seed(123)
lasso_tfidf <- cv.glmnet(x=dfmat_train_tfidf,
                   y=train$category,
                   family="binomial", 
                   alpha=1,  # alpha = 1: LASSO
                   parallel=TRUE, nfolds = 10,
                   intercept=TRUE)

Sys.time() -x 
lasso
lasso_tfidf

```

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T, preview=TRUE}
# Plot lasso without cv and with cv to mark lamda.min and lamda.1se

par(mfrow=c(1,2))
plot(lasso_1, xvar="lambda", main="Lasso penalty \n\n")
abline(v=log(lasso$lambda.min), col="red", lty="dashed")
abline(v=log(lasso$lambda.1se), col="blue", lty="dashed")

plot(lasso_1_tfidf, xvar="lambda", main="Lasso penalty tfidf \n\n")
abline(v=log(lasso_tfidf$lambda.min), col="red", lty="dashed")
abline(v=log(lasso_tfidf$lambda.1se), col="blue", lty="dashed")

```

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
par(mfrow=c(1,2))
plot(lasso,main="Lasso penalty\n\n")
plot(lasso_tfidf,main="Lasso penalty tfidf\n\n")

```

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
# Predict 
lasso
lasso_tfidf
lasso$lambda.1se
lasso$lambda.min
actual_class <- as.factor(test$category)
predicted_class.lasso <- predict(lasso, newx=dfmat_matched,s="lambda.min", type="class")

tab_class.lasso <- table(predicted_class.lasso, actual_class)

confusion_matrix_lasso <- confusionMatrix(tab_class.lasso, mode="everything", positive="covid")


##tfidf
lasso_tfidf$lambda.1se
lasso_tfidf$lambda.min
actual_class_tfidf <- as.factor(test$category)
predicted_class.lasso_tfidf <- predict(lasso_tfidf,
                                       newx=dfmat_matched_tfidf,s="lambda.min",
                                       type="class")

tab_class.lasso_tfidf <- table(predicted_class.lasso_tfidf, actual_class)

confusion_matrix_lasso_tfidf <- confusionMatrix(tab_class.lasso_tfidf, mode="everything", positive="covid")

confusion_matrix_lasso
confusion_matrix_lasso_tfidf
```

The best model in the end retains 65 variables. If we used tf-idf weighting, the best model retains 69 of the 1219 variables. That is great way to reduce the irrelevant features. We can have a look at some of these variables using the `vip`package. VIP ranks based on their importance scores.

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
library(vip)
vip(lasso_tfidf, 10)
```

Let's see if this this variables are also the main features in the ridge model.

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
vip(ridge_min_tfidf, 10)
```

# Elastic net

In my experience with text data, I found elastic net regression having problems with the matrix format of the dfm objects resulted from quanteda. I will have to convert the matrix formats to data frames.

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
da_train <- cbind(category=train$category, 
                  convert(dfmat_train, to="data.frame")) 

da_train_tfidf <- cbind(category=train$category, 
                        convert(dfmat_train_tfidf, to="data.frame")) 
```

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
ncol(dfmat_train)
ncol(da_train)

da_train <- da_train[,-2] #the document identifier variable should be removed
ncol(da_train)

## The tfidf version 
ncol(dfmat_train_tfidf)
ncol(da_train_tfidf)

da_train_tfidf <- da_train_tfidf[,-2] #the document identifier variable should be removed
ncol(da_train_tfidf)

```

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
da_train_xmatrix <- da_train[,-1]  %>% as.data.frame() %>% as.matrix() 
da_train_xdf <- da_train  %>% as.data.frame()

#for the tf-idf pre-processed data 

da_train_xmatrix_tfidf <- da_train_tfidf[,-1]  %>% as.data.frame() %>% as.matrix() 

da_train_xdf_tfidf <- da_train_tfidf  %>% as.data.frame()

```

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
da_test_match <- cbind(category=test$category, convert(dfmat_matched, to="data.frame")) 

da_test_match <- da_test_match[,-2] 
ncol(da_test_match)
da_test_xmatrix <- da_test_match[,-1]  %>% as.data.frame() %>% as.matrix() 
ncol(dfmat_matched)
ncol(da_test_xmatrix)

# Do the same for the tfidf data

da_test_match_tfidf <- cbind(category=test$category, convert(dfmat_matched_tfidf, to="data.frame")) 

da_test_match_tfidf <- da_test_match_tfidf[,-2] #the document identifier variable should be removed
ncol(da_test_match_tfidf)

da_test_xmatrix_tfidf <- da_test_match_tfidf[,-1]  %>% as.data.frame() %>% as.matrix() # remove the dependent variable

ncol(dfmat_matched_tfidf)
ncol(da_test_xmatrix_tfidf)
```

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
# Fit elastic net regression with 10 different alpha values from 0 to 1

x <- Sys.time()

set.seed(223)
y=ifelse(da_train_xdf$category=="covid", "1", "0") # convert to numeric labels
                       
cv_glmnet_10_roc <- train(x = da_train_xdf[,-1], 
                          y = y, type.measure="auc", method="glmnet",
                          family="binomial", 
                          traControl=trainControl(method="cv", number=10),
                          parallel=TRUE,
                          tuneLength=10) # I will use 10 different alpha values between 0 and 1

x-Sys.time()

#tfidf
x <- Sys.time()

set.seed(223)
cv_glmnet_10_roc_tfidf <- train(x = da_train_xdf_tfidf[,-1], 
                                y = y, type.measure="auc", method="glmnet",
                                family="binomial",
                              traControl=trainControl(method="cv",number=10), 
                                parallel=TRUE,
                                tuneLength=10) 

x-Sys.time()

```

Let's visualize the two models

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
library(ggplot2)

ggplot(cv_glmnet_10_roc)
#Tf-idf
ggplot(cv_glmnet_10_roc_tfidf)

```

```{r echo=T, message=FALSE, warning=FALSE, cache=TRUE, dpi=150, include=T}
## Predict using the belastic model cv_glmnet_50
predicted_class.elastic_10 <- predict(cv_glmnet_10_roc, 
                                      da_test_xmatrix, 
                                      cv_glmnet_10_roc$lamda.min, type="raw")

predicted_class.elastic_10 <- as.factor(ifelse(predicted_class.elastic_10==0, 
                                               "sars", "covid"))

confusion_mat_elastic_net <- confusionMatrix(predicted_class.elastic_10, 
                                      actual_class, mode="everything",
                                      positive="covid") 

#Predict the tfidf weighted data

predicted_class.elastic_10_tfidf <- predict(cv_glmnet_10_roc_tfidf,
                                            da_test_xmatrix_tfidf,
                                            cv_glmnet_10_tfidf$lamda.min, 
                                            type="raw")

predicted_class.elastic_10_tfidf <- as.factor(ifelse(predicted_class.elastic_10_tfidf==0, "sars", "covid"))
confusion_mat_elastic_net_tfidf <- confusionMatrix(predicted_class.elastic_10_tfidf, 
                                            actual_class, 
                                            mode="everything", positive="covid") 

confusion_mat_elastic_net
confusion_mat_elastic_net_tfidf

```

# Final remarks

Notice, Lasso and Elastic net models gave us superior prediction performances. A model with a sensitivity of 99%, and a specificity of 88%, and precision of more than 90% is extraordinarily superior to me! Sensitivity and specificity are not affected by prevalence. But, precision(positive predictive value) and negative predictive values are influenced by prevalence.It is possible to calculate confidence intervals for sensitivity and specificity. But, I will not do that here. Since the prevalence of covid abstracts is high(52%), the precission of my predictive models are all high. This may not be the case if your model is dealing with rare cases.

[This book](https://bradleyboehmke.github.io/HOML/) by *Bradley Boehmke & Brandon Greenwell is a good reference to learn about machine learning*. It is freely available. But, you can also buy a hard copy. It is one of my favorite machine learning books.

# Contact

[\@MihiretuKebede1](https://twitter.com/MihiretuKebede1)
